### 2020-10-07 paper.Rmd
more work on formalization

$$
\bigcup\limits_{k = 1}^{K} C_j = X
$$
$$
C_j \cap C_k = \emptyset, \; \forall j, k \in \{1, \ldots, K\}, \; j \ne k
$$

@@

For a data matrix $\mathbf{X}_{N \times P} = \{x_{ij}\}_{N \times P}$, k-means 
clustering aims to minimize the error sum of squares (SSE), also known as the 
within-group variance [@spath1986]. The SSE is the sum of the squared Euclidean 
distances between each data point and its cluster centroid; the *k*'th centroid 
is a vector of length $P$, where each entry is the mean value of one of the $P$ 
attributes across the observations belonging the *k*'th cluster ($k = 1, \ldots, 
K$). The centroid of the *k*'th cluster has the notation 
$\overline{\mathbf{x}}^{(k)}$ with
$$
\overline{\mathbf{x}}^{(k)} = 
  (\overline{x}_1^{(k)}, \overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})
$$

The mean $\overline{x}_j ^{(k)}$ of the *j*th variable on the *k*th cluster is 
computed as:

@@

Based on the original data and the cluster centroids, the SSE---i.e., the k-means 
objective function---is given as:

@@

K-means anticlustering---i.e., maximizing the SSE---minimizes differences 
between the means of the input variables. In the following, a rather 
straight-forward extension of the SSE criterion is formulated that can be used 
to minimize differences with regard to the variances of the input variables at 
the same time.

For the sake of simplicity, we first consider a set of $N$ one-dimensional 
observations $(x_1, \ldots, x_N)$. The extension to the general case of $P$ 
attributes will be straight forward, because the formulation of the new 
criterion effectively reduces to the generation of new variables, leaving the 
k-means criterion SSE per sé unchanged.

The sample variance of this set of 
observations is computed as the mean of the squared distances between 
observations and their mean $\mathbf{x}$:

@@

Note at this point that the sample variance is basically defined as a mean 
value. We can use this insight and employ k-means anticlustering to equalize 
this *mean* across groups. For a given $K$, groups can be created that are 
similar with regard to the means of $y$---i.e., the variance of $x$---by 
maximizing the SSE, in this one-dimensional case given as:

$$
\mathit{SSE} = 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{i} - \overline{y} ^{(k)})^2
$$

## Weighted-sum approach

$$
\mathit{SSE_{comb}} = 
  w_1
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (x_{i} - \overline{x} ^{(k)})^2 + 
  w_2
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (y_{i} - \overline{y} ^{(k)})^2
$$

@@

$$
\mathit{SSE_{comb}} = 
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (x_{i} - \overline{x} ^{(k)})^2 + 
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (y_{i} - \overline{y} ^{(k)})^2 \\
  = \sum\limits_{k=1}^{K} 
     \sum\limits_{i \in C_k}
     ((x_{i} - \overline{x} ^{(k)})^2 + 
     (y_{i} - \overline{y} ^{(k)})^2)
$$

When defining $z = (x, y)$, the SSE reduces to the standard case in (X):

$$
\mathit{SSE} = 
  \sum\limits_{j=1}^{2} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (z_{ij} - \overline{z}_j ^{(k)})^2
$$

>>>>

$$
\bigcup\limits_{k = 1}^{K} C_j = X
$$

$$
C_j \cap C_k = \emptyset, \; \forall j, k \in \{1, \ldots, K\}, \; j \ne k
$$

@@

For a data matrix $\mathbf{X}_{N \times P} = \{x_{ij}\}_{N \times P}$, k-means 
clustering aims to minimize the error sum of squares (SSE), also known as the 
within-group variance [@spath1986]. The SSE is the sum of the squared Euclidean 
distances between each data point and its cluster centroid. The *k*'th centroid 
is a vector of length $P$, where each entry is the mean value of one of the $P$ 
attributes, computed across the observations belonging the *k*'th cluster ($k = 
1, \ldots, K$). The centroid of the *k*'th cluster $\overline{\mathbf{x}}^{(k)} 
= (\overline{x}_1^{(k)}, \overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})$ is 
computed as:

@@

Based on the data matrix $\mathbf{X}_{N \times P} = \{x_{ij}\}_{N \times P}$, 
a partitioning of the data into $K$ clusters, the cluster centroids, the 
SSE---i.e., the k-means objective function---is then given as:

@@

K-means anticlustering---i.e., maximizing the SSE---minimizes differences 
between the means of the input variables. In the following, a rather 
straight-forward bicriterion extension of the SSE criterion is formulated that 
can be used to minimize differences with regard to the variances of the input 
variables at the same time.

For the sake of simplicity, we first consider a set of one-dimensional 
observations $(x_1, \ldots, x_N)$ to motivate the new criterion. The extension 
to the general case of $P$ attributes will be straight forward, because the 
formulation effectively reduces to the generation of new variables, whereas the 
original k-means criterion remains unchanged.

The sample variance of a set of observations $x = (x_1, \ldots, x_N)$ is 
computed as the mean of the squared distances between observations and their 
mean $\mathbf{x}$:

@@

We can use the insight that the sample variance is basically defined as an 
average, implying that k-means anticlustering can be employed to equalize this 
average across different groups. This is the central insight leading to the new 
criterion. In the one-dimensional case as discussed here, the SSE that 
represents discrepancies with regard to the variance, is given as:

$$
\mathit{SSE_{Var}} = 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{i} - \overline{y} ^{(k)})^2
$$

This criterion, reflecting discrepancies in variance across groups, is extended 
to the general case of $P$ attributes as

$$
\mathit{SSE_{Var}} = 
  \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{ij} - \overline{y}_j ^{(k)})^2
$$

with

$$
\overline{y}_j ^{(k)} = \frac{1}{n_k} \sum\limits_{i \in C_k} y_{ij}
$$

and

$$
y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2.
$$

## Bicriterion formulation

It will be rarely of interest to maximize $SSE_{Var}$ (equalizing variances) 
without also considering the standard $SSE$ (i.e., equalizing means). A 
combination of $SSE_{Var}$ and $SSE$, however, can be employed to simultaneously 
optimize both objectives. That is, formulating a criterion incorporating both 
discrepancies in means as well as variances to minimize difference with regard 
to either. The most simple approach to such a bicriterion optimization is the 
weighted sum approach. That is,

Combining (X) and (Y), we can obtain a combined criterion formulation as a 
weighted sum criterion.

$$
\mathit{SSE_{bicriterion}} = 
  w_1
  \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x_{ij} - \overline{x}_j ^{(k)})^2 + 
  w_2
  \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{ij} - \overline{y}_j ^{(k)})^2
$$

@@

$$
\mathit{SSE_{bicriterion}} = 
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (x_{i} - \overline{x} ^{(k)})^2 + 
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (y_{i} - \overline{y} ^{(k)})^2 \\
  = \sum\limits_{k=1}^{K} 
     \sum\limits_{i \in C_k}
     ((x_{i} - \overline{x} ^{(k)})^2 + 
     (y_{i} - \overline{y} ^{(k)})^2)
$$

When defining $z_i = (z_{i1}, z_{i2})$ with $z_{i1} = x_i$ and $z_{i2} = y_i$, 
the bicriterion SSE is computed as

$$
\mathit{SSE_{bicriterion}} =
  \sum\limits_{j=1}^{2} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (z_{ij} - \overline{z}_j ^{(k)})^2
$$
